---
title: Fun With Spotify's API
author: richjand
date: '2021-07-05'
slug: fun-with-spotify-s-api
categories: []
tags: []
---

```{r setup, include=FALSE}
library(nflfastR)
library(spotifyr)
library(tidyverse)
library(purrr)
library(plotly)

knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

Sys.setenv(SPOTIFY_CLIENT_ID = '8aa543e9b28d48239f55df2abc89f2aa')
Sys.setenv(SPOTIFY_CLIENT_SECRET = '0b90618c89ca4c5887ef0fd3140e204c')

```

I've been interested in getting better with purrr's map functions and have always wanted to get my hands dirty with the Spotify API so this post has a little bit of both. I'm going to use the wonderful [https://github.com/charlie86/spotifyr](spotifyr) package to take a look at trends in which NFL players are hot topics of conversation on top Fantasy Football podcasts.

The first step was to identify a list of podcasts and pull their information. Based on some lists of the most popular podcasts and my own domain knowledge I settled on the following:

  - Fantasy Footballers
  - The Ringer Fantasy Football Show
  - CBS fantasy football today
  - Fantasypros
  - Fantasy Focus Football
  - Late Round Podcast
  - Establish the Run
  - The Audible
  - Yahoo Fantasy Football Forecast
  - Rotowire
  - PFF Fantasy 
  - The Underdog Football Show
  
The first step was to pull show descriptions from the last 50 episodes of each show. This is made very easy with the <code>get_show_episides()</code> funciton from spotifyr. To clean everything and return it in exactly the form I want I wrote a little function and used <code>map_dfr</code> to apply that function for each show and return a data frame with show id, release date, and the show's text description for each episode. This left me with 582 total episodes.

```{r}
pull_descriptions <- function(id){
  show <- spotifyr::get_show_episodes(id, limit = 50)
  df <- data.frame(id = show$id, description = show$description, release_date = show$release_date)
  df$release_date <- lubridate::ymd(df$release_date)
  return(df)
}

ids <- c("5RaNsb5sKEBleahQa4MVC5",
         "0XLPhMzcKmxoNziHkVkYpR",
         "2fEvGGxwXqSM8xuSNgxjFR",
         "1YM5ymt3vWVfdHzVAEzq2w",
         "55toF30GeLKhJYGr3JPQpG",
         "3sfbS4uuJNJtUTdnBG1KkI",
         "0gzBznDnd0yGIJ1hcv2NTW",
         "70E33T64jsqzqr9V0L9CFr",
         "0wf5ZBFRJnMSwIEgzhO2MQ",
         "3yJGYiR71iW2U5oyMG3jE6",
         "5Cph8h96Td7qaxFBxPADc9",
         "4k37aIxrGzggMOIvYrWBQb")

df <- map_dfr(ids, pull_descriptions)
```

Next, I want to find which players are being talked about. To do this I pulled an up-to-date list of player names from the nflfastr package using the <code>fast_scraper_roster</code> function.

```{r}
players <- fast_scraper_roster(2021) %>%
  filter(position %in% c('QB','RB','WR','TE'))
```

We want to identify every player mentioned in the episode's description so we'll create another function and use purrr to apply that function to each podcast episode. The function will return a data frame with a row for each player matched on a given day so if the show description has four player matches the function will return 4 rows.

```{r}
get_matched_players <- function(show, date, description, player_vec){
  matches <- stringr::str_match(description, player_vec)
  matches <- matches[!is.na(matches)]
  if (length(matches) > 0){
    return_df <- data.frame(id = show, release_date = date, player = matches)
    return(return_df)
    }
}

player_matches <- pmap_dfr(.l = list(df$id, df$release_date, df$description), 
                           .f = get_matched_players, 
                           player_vec = players$full_name)

filter(player_matches, id == '6SlmIU5Vfn0j0T3K5LiPJp' & release_date == '2021-06-08')
```

At this point we have all of our matched players but we'll want to do a little more data manipulation. First, this data is very noisy day-to-day. There are certain days of the week where podcasts tend to come out and there are big seasonality effects where shows tend to release fewer episodes immediately after a season ends.

To deal with this we're going to calculate rolling averages by player and date, counting the total number of mentions in the two weeks leading up to a given date. To deal with the fact that there is more discussion generally as we get closer to the season we will define a metric, mention share, that is the rolling average of the player's mentions at a given date divided by the rolling average of total player mentions at that same date.

Before calculating rolling averages I'll want to make sure I have an entry for each player and date combination. The next three data frames we'll create are counting player mentions by day, total mentions by day, and a rolling average of total mentions over a 14 day period.

```{r}
full_mat <- expand.grid(seq(min(df$release_date), max(df$release_date), by = 1), 
                        unique(player_matches$player))
full_mat <- data.frame(full_mat)
names(full_mat) <- c('release_date','player')

player_mentions_by_day <- full_mat %>%
  left_join(player_matches, by = c('release_date','player')) %>%
  mutate(mentioned = if_else(is.na(id), 0, 1)) %>%
  group_by(release_date, player) %>%
  summarise(n_mentions = sum(mentioned)) %>%
  ungroup()

player_mentions <- full_mat %>%
  left_join(player_matches, by = c('release_date','player')) %>%
  mutate(mentioned = if_else(is.na(id), 0, 1)) %>%
  group_by(player) %>%
  summarise(player_mentions_total = sum(mentioned)) %>%
  ungroup()


total_mentions_by_day <- player_mentions_by_day %>%
  group_by(release_date) %>%
  summarise(total_mentions = sum(n_mentions)) %>%
  ungroup() %>%
  mutate(rolling_total_mentions = zoo::rollsum(total_mentions, 
                                               align = 'right', 
                                               k = 14, na.pad = T)
         )


```

I wasn't able to get player rolling averages to play nicely inside of dplyr so I created yet another function and applied over each player in the data, returning a data frame with each player's 14 day rolling average. At that point we can join our two rolling average data frames and divide the player rolling average by the total rolling average to get our mention share metric.

```{r}
split_roll <- function(player_name, df){
  dat <- filter(df, player == player_name) %>%
    mutate(rolling_player_mentions = zoo::rollsum(n_mentions, align = 'right', k = 14, na.pad = T))
  return(dat)
}

rolling_player_mentions <- map_dfr(.x = unique(player_mentions_by_day$player), 
                                   .f = split_roll, 
                                   df = player_mentions_by_day)

final_dat <- rolling_player_mentions %>%
  left_join(total_mentions_by_day, by = 'release_date') %>%
  filter(release_date > '2021-01-01') %>%
  mutate(mention_share = rolling_player_mentions/rolling_total_mentions)
```

## Visuals

At this point we have our data and can start to make some fun plots! Aaron Rodgers was in the news a ton this offseason and we can track exactly how the fantasy podcast world responded. We see a big uptick in mentions immediately after Packers were eliminated from the playoffs and Rodgers made some cryptic comments about play calling as well as big uptick in mentions around the draft in late April when some reporters started discussing him wanting to leave Green Bay.

```{r}
rodgers <- final_dat %>%
  filter(player == 'Aaron Rodgers') %>%
  ggplot(aes(x = release_date, y = mention_share)) +
  geom_line(alpha = .5) +
  geom_point(alpha = .5) +
  theme_bw() +
  xlab('Date') +
  ylab('Mention Share, 14 Day Rolling Average') +
  labs(title = 'Aaron Rodgers Offseason')
  
ggplotly(rodgers)
```

We can also look at how the focus has changed on different players at different points in the offseason. Below we're going to pull the 10 most mentioned running backs and plot their mention share across time.

```{r}
rbs <- players %>%
  filter(position == 'RB') %>%
  select(full_name, position) %>%
  inner_join(player_mentions, by = c('full_name' = 'player')) %>%
  arrange(desc(player_mentions_total)) %>%
  slice(1:10)

rb_mentions <- final_dat %>%
  inner_join(rbs, by = c('player' = 'full_name'))

top_10_rb <- rb_mentions %>%
  ggplot(aes(x = release_date, y = mention_share)) +
  geom_line(alpha = .5) +
  geom_point(alpha = .5) +
  theme_bw() +
  xlab('Date') +
  ylab('Mention Share') +
  facet_wrap(~player)

ggplotly(top_10_rb)
```
